---
title: "Data Analysis Project"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = "")
```



# Student Information

**Name**: Nandha Kishore Sangem
**G Number**:G01280660




# Introduction

This semester we will be working with a dataset of all domestic outbound flights from Dulles International Airport in 2016.

Airports depend on accurate flight departure and arrival estimates to maintain operations, profitability, customer satisfaction, and compliance with state and federal laws. Flight performance, including departure and arrival delays must be monitored, submitted to the Federal Aviation Agency (FAA) on a regular basis, and minimized to maintain airport operations. **The FAA considered a flight to be delayed if it has an arrival delay of at least 15 minutes**.

The executives at Dulles International Airport have hired you as a Data Science consultant to perform an exploratory data analysis on all domestic flights from 2016 and produce an executive summary of your key insights and recommendations to the executive team.

Before you begin, take a moment to read through the following airline flight terminology to familiarize yourself with the industry: [Airline Flight Terms](https://www.myairlineflight.com)



## Dulles Flights Data

The `flights_df` data frame is loaded below and consists of 33,433 flights from IAD (Dulles International) in 2016. The rows in this data frame represent a single flight with all of the associated features that are displayed in the table below.

**Note**: If you have not installed the `tidyverse` package, please do so by going to the `Packages` tab in the lower right section of RStudio, select the `Install` button and type `tidyverse` into the prompt. If you cannot load the data, then try downloading the latest version of R (at least 4.0). The `readRDS()` function has different behavior in older versions of `R` and may cause loading issues.



```{r}

library(tidyverse)
library(ggplot2)
library(plotly)
flights_df <- readRDS(url('https://gmubusinessanalytics.netlify.app/data/dulles_flights.rds'))

```


### Raw Data

```{r}

flights_df

```


# Exploratory Data Analysis


Executives at this company have hired you as a data science consultant to evaluate their flight data and make recommendations on flight operations and strategies for minimizing flight delays.

You must think of **at least 8 relevant questions** that will provide evidence for your recommendations.

The goal of your analysis should be discovering which variables drive the differences between flights that are early/on-time vs. flights that are delayed.

Some of the many questions you can explore include:

-   Are flight delays affected by taxi-out time? Do certain airlines or time of 
    year lead to greater taxi out times (i.e. traffic jams on the runways)?

-   Are certain times of the day or year problematic?

-   Are certain destination or airlines prone to delays?


You must answer each question and provide supporting data summaries with either a summary data frame (using `dplyr`/`tidyr`) or a plot (using `ggplot`) or both.

In total, you must have a minimum of 5 plots and 4 summary data frames for the exploratory data analysis section. Among the plots you produce, you must have at least 4 different types (ex. box plot, bar chart,
histogram, heat map, etc...)

Each question must be answered with supporting evidence from your tables and plots.


## Question 1



**Question**:

how the month of the year affects the departure delay

**Answer**:
From this data we can see that months June and July have affected departure delay for airlines. Skywest airlines got affected the most in feb and June months whereas Frontier airlines got affected the most in July, August and september month 


To add additional R code chunks for your work, select `Insert` then `R` from the top of this notebook file.

```{r}


flights_df %>%
  group_by(month) %>%
  summarise(AvgDelay=mean(dep_delay))

ggplot(flights_df, aes(x=factor(month), y=dep_delay,fill=airline)) +stat_summary(fun.y="mean", geom="bar")+
xlab("Months") +
ylab("Departure delay of the flight ") +
ggtitle("Departure delay of flights in all months")+
theme(plot.title =element_text(hjust = 0.5),axis.text.x = element_text(size=8,angle=90))

```



## Question 2



**Question**:

Total number of flights operating for each airline from Dulles International airport in 2016

**Answer**:
There are total 33433 flights operated at Dulles international airport out of which 61.77% of flights(20653) are from United airline and alaska airlines contribute least percentage of flights 1.07%



```{r}
f1<-flights_df %>%
  count(airline,total_flights=n())%>%
  mutate(Percentage=(n/total_flights)*100)
f1
ggplot(data=f1,aes(x=airline,y=Percentage))+geom_point()
ggplot(data=f1, aes(x=airline, y=Percentage,fill=airline)) +stat_summary(fun.y="mean", geom="bar")

```



## Question 3



**Question**:

How the season affects the departure delay

**Answer**:

From below summary, Summer has the maximum mean of departure delay compared to all other seasons with 15.57. As most of the filghts operating in June, July months we can observe departure delay more in summer season and moreover United airlines is the one which has more number of flights operating in all the seasons with high departure delay time.


```{r}

  
flights_df %>%
  group_by(season) %>%
  summarise(AvgDelay=mean(dep_delay))

ggplot(flights_df, aes(season, dep_delay, fill=airline))+ geom_col()+
labs(title = "Effect of season on Departure Delay")+xlab("Season") +ylab ("Departure Delay")+theme(plot.title =element_text(hjust = 0.5))

```




## Question 4



**Question**:

To which regions do flights Depart late from the Dulles International Airport?

**Answer**:

Here we can see flights have high departure time delay for regions Midwest and Southwest from IAD when compared to other regions. In Midwest region departure time of flights crossed even 1200 mins. Also in regions West and South we can see some flights which have departed early before the scheduled time.


```{r}

ggplot(data = flights_df,aes(x = dest_airport_region, y = dep_delay)) + geom_boxplot()+
xlab("Region") +
ylab("Departure delay of the flight \n in minutes") +
ggtitle("Departure delay of flights from Dulles International Airport")+
theme(plot.title =element_text(hjust = 0.5))+
  scale_x_discrete(guide = guide_axis(angle=90))




```




## Question 5



**Question**:

Is there a specific airline with a high frequency of flight delays?

**Answer**:

There are 1982 United airline flights which have more than 30 min of departure delay and Alaska airlines with least number of flights which is 37. From the data we can see Skywest airline has 1244 mins of departure delay and 1228 mins of arrival delay which is highest compared to all the airlines in year 2016.


```{r}

flights_df%>%
  select(airline,arrival_delay,dep_delay)%>%
  filter(dep_delay>30)%>%
  arrange(desc(dep_delay))%>%
  count(airline)

p<-flights_df %>% ggplot(aes(x=  dep_delay, y =  arrival_delay,color=airline)) +geom_jitter()+
  labs(title = "Airlines that have number of delays")+
  xlab("Departure delay")+
  ylab("Arrival Delay")+theme(plot.title = element_text(hjust = 0.5))
ggplotly(p)

```


## Question 6



**Question**:

which month has most number of flights operating to and from Dulles International airport

**Answer**:
United airlines is the one airline carrier which has most number of flights operating in every month of year 2016 whereas for all other airlines flights operating are less than 500 in every month. For United airlines highest number of flights(1994) are operated in July month compared to other months.  


```{r}

flights_df1<-flights_df %>%
  select(month,airline) %>%
  group_by(month) %>%
  count(airline)
flights_df1
ggplot(data = flights_df1 ) + geom_point(aes(x = month, y = n,color=airline))

```


## Question 7



**Question**:

How many flights were scheduled to depart before 6 a.m.?

**Answer**:

According to the data, the months of June and July had the highest number of early flights. In July, 94 flights took off before 6a.m., whereas 91 flights took off before 5 a.m. in June. In addition, 44 flights left in late July.


```{r}

flights_df %>%
    group_by(month) %>%
    summarize(early_flight = sum(dep_time<6.00),late_flight=sum(dep_time>23.00))


```



## Question 8



**Question**:

How's departure delay density of all flights at IAD in 2016

**Answer**:

We can see high density of the flights for departing are over and before the 0 line which indicates that most of flights are departing on time without any delay and there are some flights with delay less than 100 mins which can be depicted from below visualization.


```{r}

ggplot(flights_df, aes(dep_delay)) + geom_density() + xlim(-100,200)+
xlab('Departure Delay in minutes') + ylab('Density of Departure Delay')

ggplot(data=flights_df,aes(x=dep_delay,fill=airline))+geom_histogram()+
  labs(title = "Arrival Time Delays in Minutes")+
  xlab("Arrival delay")+
  ylab("Frequency")

```



## Question 9



**Question**:

How's arrival delay frequency of all the flights in 2016 at IAD

**Answer**:

From below histogram we can see arrival delay frequencies of flights at IAD in 2016.Most of the flights are arriving on time and some are arriving before scheduled arrival time.we can observe arrival delay frequency is less for flights with increase in arrival delay time.


```{r}

ggplot(data=flights_df,aes(x=arrival_delay,fill=airline))+geom_histogram()+
  labs(title = "Arrival Time Delays in Minutes")+
  xlab("Arrival delay")+
  ylab("Frequency")

```


# Summary of Results

Write an executive summary of your overall findings and recommendations to the executives at Dulles Airport. Think of this section as your closing remarks of a presentation, where you summarize your key findings and make recommendations on flight operations and strategies for minimizing flight delays.

Your executive summary must be written in a [professional tone](https://www.universalclass.com/articles/writing/business-writing/appropriate-tone-in-business-communications.htm), with minimal grammatical errors, and should include the following
sections:

1.  An introduction where you explain the business problem and goals of your data analysis

    -   What problem(s) is this company trying to solve? Why are they important
        to their future success?

    -   What was the goal of your analysis? What questions were you 
        trying to answer and why do they matter?


2.  Highlights and key findings from your Exploratory Data Analysis section

    -   What were the interesting findings from your analysis and
        **why are they important for the business**?

    -   This section is meant to **establish the need for your recommendations** 
        in the following section
        

3.  Your recommendations to the company

    -   Each recommendation must be supported by your data analysis results

    -   You must clearly explain **why** you are making each recommendation and which results 
        from your data analysis support this recommendation

    -   You must also describe the potential business impact of your recommendation:

        -   Why is this a good recommendation?

        -   What benefits will the business achieve?
      


## Executive Summary

Please write your executive summary below. If you prefer, you can type your summary in a text editor, such as Microsoft Word, and paste your final text here.

**Introduction**
Problem Statement
One of the most typical issues that makes airport authorities extremely difficult at times is flight delays. In several situations, airport officials have been proven to be accountable for work mismanagement and failure to take appropriate actions, which has caused many significant losses for airlines.

In order to overcome this issues we need to go through all the data provided by airport authorities and analyze the data and find solutions for these issues. Along with this we need to provide respective stats and visualizations of what factors are causing these issues to make it understand clearly for authorities.

**Key Findings**

There are total of 33433 flights operated from IAD in 2016. From this United airlines has operated most number of flights with 20653 which comprises almost 62% of flights. As United airlines has maximum share for operating flights, most of arrival delays and departure delays are caused by these airlines compared to other airlines

We can see that most number of flights have operated in June, July months. As frequency is high in these months average departure delay is also high for these months which is considered as Summer season. So Summer is considered as a busy season among all the seasons

From our visualizations we can observe a direct relation for arrival delay and departure delay. with increase in departure delay there is an increase in arrival delay. For example Skywest airline has 1244 mins of departure delay and 1228 mins of arrival delay which is highest compared to all the airlines in year 2016.

**Recommendations** 
As there is high frequency of delays in summer season it would be best for airport authorities to manage work effectively and connect with airlines to ensure good quality time for all the passengers on their journeys.

Authorities need to look into flights departing to Midwest and Southwest regions as flights have more departure delay time operating to these regions.

Also from the stats as United airlines is the carrier which operates high number of flights and causing more arrival and departure delays compared to other airlines, authorities should work with this airlines and provide extra support to make sure follow the scheduled times and notify passengers before time in case of any delays.

From the insights, these are some of possible recommendation for the data provided for smooth and proper functioning of airport. There is need of extra information such as a flight's airtime, weather prediction, temperature and other parameters which are critical for gaining more insights and making good suggestions.









